---
title: "static-mini1"
format: pdf
author: Morgan Ryan
editor_options: 
  chunk_output_type: console
---

[Link To Interactive Plots](https://github.com/morgan-ryan15/SDS264_F24/blob/main/SDS264/projects/interactive-mini1.qmd)


```{r}
#| message: false
#| warning: false
library(tidyverse)
library(maps)
library(viridis)
library(janitor) # r tip of day??
library(leaflet)
library(htmltools)
library(glue)
library(sf)
#data for proj
death_data <- read_csv("~/SDS264_F24/SDS264/Data/death.csv")

#data for map
us_states <- map_data("state")
head(us_states)

#sf 
states <- read_sf("https://rstudio.github.io/leaflet/json/us-states.geojson")  #<2>
class(states) #<3>
states

```

```{r}
#examine state levels
unique(us_states$region)
unique(death_data$State)
```

```{r}
#Graph #1: Static with Numeric Value
num_death_data <- death_data |>
  
  #clean column names and select cols
  clean_names() |>
  select(year, cause_name, state, deaths) |>
  mutate(state = str_to_lower(state)) |>
  
  #filter -> wanted data to avoid many-to-many issue before join by shrinking dataset down
  filter(!(state %in% c("alaska", "hawaii", "united states"))) |>
  filter(cause_name %in% c("Heart disease"), year == 2017) |>
  

  #join data from us states into numerical death data set
  right_join(us_states, by = c("state" = "region")) |>
  rename(region = state) 

  #plotting data 
num_death_data |>
  ggplot(mapping = aes(x = long, y = lat,
                       group = group)) + 
  geom_polygon(aes(fill = deaths), color = "black") +
  coord_map() +  
  theme_void() +  
  scale_fill_viridis() +
  labs(   
    title = "United States: Deaths By Heart Disease in 2017",
    fill = "Number of Deaths",
    caption = "Data Source: data.gov")
```
Description Graph #1:
This is a static graph with a numeric variable used to fill in the United States. Using data from data.gov, specifically from 2017 focusing on the amount of deaths due to heart disease. This is an informative map, as it shows which states have had the highest and lowest number of death by heart disease. The darker the state, the fewer deaths and the brighter the state, the greater amount of deaths. 

Alt-Text Graph #1:
This is a static graph with the numeric variable "deaths" filling in the map of the United States. Deaths stands for the number of deaths recorded per each state in 2017. It ranges from 0 to just over 60000 with colors associated being darker (purple) for lower numbers and brighter (yellow) for higher numbers of deaths. By viewing the appearance, one evident takeaway from the graph is California having the highest numbers of death due to heart disease in 2017. 


```{r}
#Graph #2: Static with categorical value

#Create data set for categorical var
cat_death_data <- death_data |>
  
  #cleaning col names and selecting cols
  clean_names() |>
  select(year, cause_name, state, deaths) |>
  mutate(state = str_to_lower(state)) |>
  
  #filtering cols for join
  filter(!(state %in% c("alaska", "hawaii", "united states"))) |>
  filter(cause_name %in% c("Alzheimer's disease", "Stroke"), year == 2017) |>
  
  #fixing two rows per state 
  pivot_wider(
    names_from = cause_name,
    values_from = deaths 
  ) |>
  
  #get counts and condition for cat var 
  clean_names() |>
  count(state, stroke, alzheimers_disease) |>
  mutate(death_type = ifelse(stroke > alzheimers_disease, "S", "A")) |>
  
  #Join data set to us_states
  right_join(us_states, by = c("state" = "region")) |>
  rename(region = state) 
  
  #plot
  cat_death_data |>
  ggplot(mapping = aes(x = long, y = lat,
                          group = group)) + 
  geom_polygon(aes(fill = death_type), color = "gray33", linewidth = 0.2) + 
  coord_map() + 
  theme_void() +  
  scale_fill_manual(values = c("plum4", "darkseagreen3"),
                    labels = c("Alzheimers", "Stroke")) +
  labs(fill = "Cause of Death",
       title = "United States: Highest Cause of Death Between Alzheimers and Stroke in 2017",
       caption = "Data Source: data.gov") 

```

Description #2:
This is a static categorical map of the United States and sourced from data.gov. Each state compares the number of deaths due to Alzheimer's and Stroke in 2017, whichever cause of death is recorded most per state is what that state is labelled. For example, Minnesota has more recorded deaths due to Alzheimer's then Strokes, so it is filled with purple to represent Alzheimer's. We can see that more states appear to have more deaths due to Stroke than Alzheimer's.
